---
title: "eBirdst Script for Chapter 1"
author: "Andres N. Rosales"
date: "2024-10-10"
output:
  word_document:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    highlight: espresso
    toc: yes
    theme: sandstone
---

# Analyzing Landscape-Level Factors Influencing Upland Sandpiper Population Dynamics

This code extracts and manipulates eBird status and trends data for the first chapter of my thesis. More infomation and visualizations can be found at <https://science.ebird.org/en/status-and-trends/species/uplsan/trends-map>

```{r}
#set wd to github folder, only for using on ECCC computer
#setwd("C:/Users/ROSALESA/Desktop/Analysis/UPSA_agri/Scripts/ebirdst")

```


## Library

```{r, message=FALSE, warning=FALSE}


library(dplyr)
library(purrr)
library(ggplot2)
library(rnaturalearth)
library(sf)
library(raster)
library(terra)
library(tinytex)
library(rmarkdown)
library(knitr)
library(ebirdst)
library(exactextractr)
library(landscapemetrics)
library(readr)
library(stringr)
library(tidyr)
library(units)
library(viridis)


```


## Extract Status and Trend Data
This code views and extracts available status and trend data for Upland Sandpiper.


Set unique key to access eBirdst data (expires Jan 18th 2025)
```{r}
set_ebirdst_access_key("suvkk96q5mrj", overwrite = TRUE)

```

Show all species with status and trend data available. We will subset to show Upland Sandpiper data availability. 
```{r}

trend_runs <- ebirdst_runs %>%
  filter(has_trends) %>%
  dplyr::select(
    species_code, common_name, trends_season, trends_region,
    trends_start_year, trends_end_year, trends_start_date,
    trends_end_date, rsquared, beta0
  )

u_runs <- trend_runs %>%
  filter(common_name == "Upland Sandpiper") %>%
  dplyr::select(trends_start_year, trends_end_year, 
                trends_start_date, trends_end_date)
```




Create an object of Upland Sandpiper trends (2012-2022). 
*waiting for new eBird data release in November
```{r}
ebirdst_download_trends("Upland Sandpiper")

utrends_st <- load_trends("Upland Sandpiper")


# Write a csv of each trend pixel 
write.csv(utrends_st, file = "Data/utrends20122022.csv", row.names = FALSE) 
```


## Vector of Spatial Trends and Abundance
load vector of the pixel points 

```{r}
utrends_sf <- st_as_sf(utrends_st, 
                       coords = c("longitude", "latitude"), 
                       crs = st_crs(4326))                   #set WGS84 as crs

#Save as a GeoPackage to working directory
write_sf(utrends_sf,"Trends_sf/utrends_sf2022.gpkg", layer = "uppies_trends")
```


#Spatial Buffer for predictor grid
Removed Alaska / Yukon points from utrends_st
```{r}

utrends_st <- utrends_st %>% filter(!(srd_id >= 126707 & srd_id <= 171704))

```


Using utrend_st make an 'sf' object. Set crs as WGS84 then reproject to NAD83 (5070)
```{r}
#set to 4326 then trnasfomr later during lsm
utrends_buff <- utrends_st |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) |>
  st_buffer(dist = set_units(13.5, "km"))                                        # Create 13.5 km buffers

ggplot() +
  geom_sf(data = utrends_buff, fill = "lightblue", color = "darkblue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "13.5 km Circular Buffers", x = "Longitude", y = "Latitude")


write_sf(utrends_buff,"C:/Users/Andres/OneDrive - University of Saskatchewan/03_Arcpro/Chapter1/Shapefiles/ebirdst/Buffer/utrends_buff.gpkg",
         layer = "ubuff")

```
upsa trends for range wide (excluding YT and AK) that are only >80% confidence
```{r}
rangewide80 <- subset(utrends_buff, abd_ppy_nonzero == "TRUE")

ggplot() +
  geom_sf(data = rangewide80, fill = "lightblue", color = "darkblue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "13.5 km Circular Buffers", x = "Longitude", y = "Latitude")

```



## Spatial Buffer for PHJV
```{r}
phjv_sf <- st_read("Trends_sf/utrends_points_phjv.shp") #shapefile of points in NAD83

phjv_buff <- phjv_sf |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) |>
  st_buffer(dist = set_units(13.5, "km")) 


ggplot() +
  geom_sf(data = phjv_buff, fill = "lightgreen", color = "darkgreen", alpha = 0.5) +
  theme_minimal() +
  labs(title = "13.5 km Circular Buffers", x = "Longitude", y = "Latitude")

#save to onedrive
write_sf(phjv_buff,"C:/Users/Andres/OneDrive - University of Saskatchewan/03_Arcpro/Chapter1/Shapefiles/ebirdst/Buffer/utrd_buff_PHJV.gpkg",
         layer = "ubuffPHJV")

#save to WD. THis is the buffer I will use for PHJV analysis and for GEOG project
write_sf(phjv_buff,"Predictor_buffer/utrd_buff_PHJV.gpkg",
         layer = "ubuffPHJV")
```


## Rasterize Upland Sandpiper trends

rasterize cumulative trend estimate using `terra` package
```{r, include=FALSE}
utrends_ras <- rasterize_trends(utrends_st, 
                                  layers = c("abd_trend"))


```


```{r}
#save as a geotiff for use in arcpro
writeRaster(utrends_ras, filename = "Rasters/utrends_ras2022.tif", overwrite = TRUE)

plot(utrends_ras)

```



#Status Predictive Importance
Downloads and views the eBird status data for Upland Sandpiper (along with PI for variables).
Only top 30 ranked PI are given in the object 'pi' These are downloaded to the eBird Directory, not in working directory
```{r}
ebirdst_download_status("uplsan", download_pis = TRUE)

upsa_pi <- list_available_pis("uplsan")
print(pi)

pi <- ebirdst_predictors
pi_des <- ebirdst_predictor_descriptions

```


## Rasterize Upland Sandpiper trends

rasterize cumulative trend estimate using `terra` package
```{r, include=FALSE}
utrends_ras <- rasterize_trends(utrends_st, 
                                  layers = c("abd_trend"))


```


```{r}
#keep projection and change in arcpro
#save as a geotiff for use in arcpro
writeRaster(utrends_ras, filename = "C:/Users/Andres/OneDrive - University of Saskatchewan/14_Arcpro/Chapter1/Shapefiles/ebirdst/utrends_ras2022.tif", overwrite = TRUE)

plot(utrends_ras)

```


## Session Info
```{r}
sessionInfo()
```

