---
title: "Rangewide Analysis"
author: "Andres N. Rosales"
date: "2025-01-24"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

# Candidate Models
**Prediction 1**

Breeding regions declining in trend will be correlated with low landscape heterogeneity and simplified agricultural landscapes.

$Trend = Δ SHDI + Δ Edge Density + Δ Cropland + (1 | Location)$


**Prediction 2**

Greater landscape heterogeneity will lower the negative effects of landscape simplification on breeding trends.
“Is the effect of landscape diversity on Upland Sandpiper population dynamics dependent on the degree of simplification?” 

$Trend = Δ SHDI + Δ Cropland + Δ SHDI * Δ Cropland + (1 | Location)$

**Prediction 3**

Forage agriculture and grasslands are associated with positive trends and high abundance. 
“How are trends in forage agriculture and grasslands associated with trends and abundance of Upland Sandpiper?”

$Trend = Δ Forage Ag + Δ Grassland + (1 | Location)$

“Does configuration of forage and grassland habitats influence their association with trends and abundance

$Trend = Δ Forage Agriculture * Δ Forage Ag LPI + Δ Grassland *	 Δ Grassland LPI + (1 | Location)$

**Prediction 4**

Trends decline or become more variable with increasing distance from the core breeding range. Conversely, trends will be more stable near the core breeding region. 

$Trend = Periphery + (1 | Location)$


**Prediction 5**

Geographical variation in abundance is driven by marginal habitat at the periphery.

$Trend = Periphery + Δ SHDI +  Δ Cropland + (1 | Location) + (1| Year)$


```{r, message=FALSE, warning=FALSE, include=FALSE}

library(dplyr)
library(purrr)
library(ggplot2)
library(rnaturalearth)
library(sf)
library(raster)
library(terra)
library(tinytex)
library(rmarkdown)
library(knitr)
library(ebirdst)
library(exactextractr)
library(landscapemetrics)
library(readr)
library(stringr)
library(tidyr)
library(units)
library(viridis)
library(progress)
library(dplyr)
library(lme4)
library(mgcv)
library(GGally)

```



```{r, include=FALSE}
all_lsm <- readRDS("Data/all_lsm.rds")
```

```{r, include=FALSE}
names(all_lsm)
```

```{r, include=FALSE}
#Clean trend data for DF
upsabuff <- readRDS("Data/upsabuff.rds")
```

```{r, include=FALSE}
#Edited trend df
upsa_trd <- upsabuff

```


```{r, include=FALSE}

upsa_trd$trend_dir <- ifelse(upsa_trd$abd_trend >0, "Increase", "Decrease")

```


# Response Distribution


Distribution of trend response variable.

```{r}
ggplot(upsabuff, aes(x = abd_trend)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  theme_minimal()

```
Almost follows a normal distribution, but without trends that are zero (stable) we observe biomodal. I could possibly look into adding a categorical variable indicating increase or decrease. I need to explore this more...



```{r}
#Distribution of abundance response variable. 
ggplot(upsabuff, aes(x = abd)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  theme_minimal()

```

Distribution of abundance estimates. 

# Predictor relationships

Correlation matric before aggregating predictors.
```{r}
library(ggcorrplot)
# List of predictors to include
predictors <- c(
   "lpi_c11_Grassland", "shdi_cNA_NA", "lpi_c09_Forage", "pland_c02_Ag", "ed_cNA_NA"
)

# Filter dataset
filtered_data <- all_lsm %>%
  dplyr::select(all_of(predictors))

# Compute correlation matrix
cor_matrix <- cor(filtered_data, use = "complete.obs", method = "pearson")

# Visualize the correlation matrix
ggcorrplot(cor_matrix, lab = TRUE, lab_size = 3, colors = c("red", "white", "blue"))
```

I still need to determine a pearsons correlation coefficient threshold to report. Some literature has used 0.7 and I will go with that for now. I still need to develop a better backing for this and if this is something I will continue to use going forward...


## Shannon Diversity Index
I want to use the slope of a linear regression for SHDI ~ year for every location as a proxy for change in SHDI. I will need to confirm SHDI changes in linearly.
fit lm and gam and compare fit for slope of SHDI as explanatory predictor
```{r}
shdi_lmm <- lmer(shdi_cNA_NA ~ year + (1|srd_id), data = all_lsm)
summary(shdi_lmm)

shdi_gamm <- gam(shdi_cNA_NA ~ s(year, k=10) + srd_id, method="REML", data = all_lsm)
summary(shdi_gamm)
AIC(shdi_lmm, shdi_gamm)
```
Big difference in AIC values when comparing GAM and lm model. the relationships of each location likely follow a linear relationship over years, allowing it to be representative of the change in diversity over time. 



Calculate change in SHDI using linear regression. THis will be used as a predictor representing the change in diversity over the trend period
```{r}
#Slope and intercept
shdi_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    shdi_slope = coef(lm(shdi_cNA_NA ~ year))[2],  
    shdi_intercept = coef(lm(shdi_cNA_NA ~ year))[1] 
  )

upsa_trd <- shdi_slopes %>%
  left_join(upsa_trd, by = c("srd_id"))
```


add mean and standard deviation of shdi for each location
```{r}
#Mean
mean_shdi <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(mean_shdi = mean(shdi_cNA_NA, na.rm = TRUE))
upsa_trd <- mean_shdi %>%
  left_join(upsa_trd, by = "srd_id")
#Standard Dev
sd_shdi <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(sd_shdi = sd(shdi_cNA_NA, na.rm = TRUE))

upsa_trd <- sd_shdi %>%
  left_join(upsa_trd, by = "srd_id")
```


```{r}
ggplot(upsa_trd, aes(x = shdi_slope, y = abd_trend, color = trend_dir)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "lm",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between SHDI Slope and Trend",
    x = "Slope of SHDI (Change over Time)",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

```

Figure reveals mostly linear with some deviation along the edges. Test lm and gam model to determine best fit.
```{r}
shdilm_model <- lm(abd_trend ~ shdi_slope, data = upsa_trd)
plot(fitted(shdilm_model), residuals(shdilm_model), main = "Residuals vs Fitted")
abline(h = 0, col = "red")
shdigam_model <- gam(abd_trend ~ s(shdi_slope), data = upsa_trd)
summary(shdilm_model)
summary(shdigam_model)

# Compare lm and gam
AIC(shdilm_model, shdigam_model)
```

Slight improvement of fit with the GAM model. However, since it is small it might be ok to assume linear relationship.





## Landscape Simplification
calculate slope of landscape simplification over 11 years
```{r}
simp_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    simp_slope = coef(lm(pland_c02_Ag ~ year))[2],  
    simp_intercept = coef(lm(pland_c02_Ag ~ year))[1] 
  )

upsa_trd <- simp_slopes %>%
  left_join(upsa_trd, by = c("srd_id"))
```

```{r}
#Mean
mean_simp <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(mean_simp = mean(pland_c02_Ag, na.rm = TRUE))
upsa_trd <- mean_simp %>%
  left_join(upsa_trd, by = "srd_id")
```

plot simplification slope against trend
```{r}
ggplot(upsa_trd, aes(x = simp_slope, y = abd_trend, color = trend_dir)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "lm",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Simplification Slope and Trend",
    x = "Slope of Simplification (Change over Time)",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```


```{r}
ggplot(upsa_trd, aes(x = mean_simp, y = abd_trend, color = trend_dir)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "lm",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Mean simplification and Trend",
    x = "Mean Simplification metric",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```



```{r}
simplm_model <- lm(abd_trend ~ simp_slope, data = upsa_trd)
plot(fitted(simplm_model), residuals(simplm_model), main = "Residuals vs Fitted")
abline(h = 0, col = "red")
simpgam_model <- gam(abd_trend ~ s(simp_slope), data = upsa_trd)
summary(simplm_model)
summary(simpgam_model)

# Compare lm and gam
AIC(simplm_model, simpgam_model)
```
## LPI Grassland
calculate the slope and average LPI for each location
```{r}
#Slope
GLPI_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    GLPI_slope = coef(lm(lpi_c11_Grassland ~ year))[2],  
    GLPI_intercept = coef(lm(lpi_c11_Grassland ~ year))[1] 
  )
upsa_trd <- GLPI_slopes %>%
  left_join(upsa_trd, by = c("srd_id"))

#Mean
mean_GLPI <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(mean_GLPI = mean(lpi_c11_Grassland, na.rm = TRUE))
upsa_trd <- mean_GLPI %>%
  left_join(upsa_trd, by = "srd_id")
```

Plot trend ~ slope and trend ~ Mean
```{r}
ggplot(upsa_trd, aes(x = GLPI_slope, y = abd_trend, color = trend_dir)) +
  geom_point(alpha = 0.3, size = 2) +  
  geom_smooth(method = "lm", se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Grassland LPI and Trend",
    x = "Slope of Grassland LPI (Change over Time)",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

ggplot(upsa_trd, aes(x = mean_GLPI, y = abd_trend, color = trend_dir)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "lm", se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Grassland LPI and Trend",
    x = "Mean Grassland LPI",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

## LPI Forage
calculate the slope and average LPI for each location
```{r}
#Slope
FLPI_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    FLPI_slope = coef(lm(lpi_c09_Forage ~ year))[2],  
    FLPI_intercept = coef(lm(lpi_c09_Forage ~ year))[1] 
  )
upsa_trd <- FLPI_slopes %>%
  left_join(upsa_trd, by = c("srd_id"))

#Mean
mean_FLPI <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(mean_FLPI = mean(lpi_c09_Forage, na.rm = TRUE))
upsa_trd <- mean_FLPI %>%
  left_join(upsa_trd, by = "srd_id")
```

Plot trend ~ slope and trend ~ Mean
```{r}
ggplot(upsa_trd, aes(x = FLPI_slope, y = abd_trend, color = trend_dir)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "loess",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Forage LPI and Trend",
    x = "Slope of Forage LPI (Change over Time)",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

ggplot(upsa_trd, aes(x = mean_FLPI, y = abd_trend, color= trend_dir)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "loess",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Forage LPI and Trend",
    x = "Mean forage LPI",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

## Edge Density
calcaulte the slope and average LPI for each location
```{r}
#Slope
ED_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    ED_slope = coef(lm(ed_cNA_NA ~ year))[2],  
    ED_intercept = coef(lm(ed_cNA_NA ~ year))[1] 
  )
upsa_trd <- ED_slopes %>%
  left_join(upsa_trd, by = c("srd_id"))

#Mean
mean_ED <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(mean_ED = mean(ed_cNA_NA, na.rm = TRUE))
upsa_trd <- mean_ED %>%
  left_join(upsa_trd, by = "srd_id")
```

Plot trend ~ slope and trend ~ Mean
```{r}
ggplot(upsa_trd, aes(x = ED_slope, y = abd_trend, color = trend_dir)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "loess", se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Edge Density and Trend",
    x = "Slope of Edge Density (Change over Time)",
    y = "Abundance Trend (abd_trend)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

ggplot(upsa_trd, aes(x = mean_ED, y = abd_trend, color = trend_dir)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "loess",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Edge Density and Trend",
    x = "Mean Edge Density",
    y = "Abundance Trend (abd_trend)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```


Proportion of grasslands
```{r}
#Slope
Gr_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    Gr_slope = coef(lm(pland_c11_Grassland ~ year))[2],  
    Gr_intercept = coef(lm(pland_c11_Grassland ~ year))[1] 
  )
upsa_trd <- Gr_slopes %>%
  left_join(upsa_trd, by = c("srd_id"))

#Mean
mean_Gr <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(mean_Gr = mean(pland_c11_Grassland, na.rm = TRUE))
upsa_trd <- mean_Gr %>%
  left_join(upsa_trd, by = "srd_id")
```


Proportion of forage
```{r}
#Slope
Fo_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    Fo_slope = coef(lm(pland_c09_Forage ~ year))[2],  
    Fo_intercept = coef(lm(pland_c09_Forage ~ year))[1] 
  )
upsa_trd <- Fo_slopes %>%
  left_join(upsa_trd, by = c("srd_id"))

#Mean
mean_Fo <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(mean_Fo = mean(pland_c09_Forage, na.rm = TRUE))
upsa_trd <- mean_Fo %>%
  left_join(upsa_trd, by = "srd_id")
```






```{r}
#save upsa_trd as rds and CSV
write_rds(upsa_trd, "Data/upsa_trd.rds")
write.csv(upsa_trd, "Data/upsa_trd.csv")
```


### Correlation matrix
matrix of relationships / distribution of response and explanatory variables after aggregating
```{r, message=FALSE}
upsa_trd$trend_dir <- ifelse(upsa_trd$abd_trend >0, "Increase", "Decrease")


agg_mean_predictors <- upsa_trd[,c( "shdi_slope", "mean_shdi",  "abd_trend", "trend_dir")]

agg_slope_predictors <- upsa_trd[,c("ED_slope","FLPI_slope", "GLPI_slope", "simp_slope",  "shdi_slope", "abd_trend", "trend_dir")]

#aggregated_predictors <- upsa_trd[,c("mean_ED","ED_slope","mean_FLPI","FLPI_slope","mean_GLPI", "GLPI_slope", "simp_slope", "mean_shdi", "shdi_slope", "abd_trend", "trend_dir")]

#Average 
ggpairs(agg_mean_predictors, ggplot2::aes(colour = trend_dir), lower = list(continuous = wrap("smooth", colour = "blue")))

#Slope
ggpairs(agg_slope_predictors, ggplot2::aes(colour = trend_dir), lower = list(continuous = wrap("smooth", colour = "blue")))

```


```{r}
ggplot(upsa_trd, aes(x = shdi_slope, y = mean_shdi, color  = trend_dir)) + 
  geom_point() + 
  geom
```



```{r}
names(upsa_trd)
```


# Model Runs
Current work in progress...

I have the data frame and predictors set up to be able to run models. 

I need to determine which predictors I will use in candidate models (mean vs slope). 
Also, I need to determine a strategy for how I will model.(How can I include abundance, spatial dependencies?, error structure?)


Model prediction 1 using gam with smoothed interaction of location
```{r}
P1gam <- gam(abd_trend ~ s(mean_shdi) + s(mean_ED) + s(mean_simp) + s(longitude, latitude), data = upsa_trd, method = "REML")
summary(P1gam)
plot(P1gam)
gam.check(P1gam)

```


Simple model examine if abundance is a significant predictor in trends 
```{r}

ggplot(upsa_trd, aes(x = abd_trend, y = abd, color = trend_dir))+ 
  geom_point()+
geom_smooth(method = "loess", se = TRUE, linetype = "dashed")

abdmodel <- gam(abd_trend ~ + s(abd) + s(longitude, latitude), data = upsa_trd, method = "REML")
summary(abdmodel)
plot(abdmodel)
```


```{r}
globalmodel <- gam(abd_trend ~ s(mean_shdi) + s(mean_ED) + s(mean_FLPI) + s(mean_GLPI) + s(simp_slope) + abd + s(longitude, latitude), data = upsa_trd, method = "REML")
summary(globalmodel)
plot(globalmodel)
gam.check(globalmodel)
```





