---
title: "rangewide-lsm"
author: "Andres N. Rosales"
date: "2025-01-13"
output:
  html_document:
    highlight: espresso
    toc: yes
    theme: sandstone
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

# Library
```{r, message=FALSE, warning=FALSE}

library(dplyr)
library(purrr)
library(ggplot2)
library(rnaturalearth)
library(sf)
library(raster)
library(terra)
library(tinytex)
library(rmarkdown)
library(knitr)
library(ebirdst)
library(exactextractr)
library(landscapemetrics)
library(readr)
library(stringr)
library(tidyr)
library(units)
library(viridis)
library(progress)
library(dplyr)
library(RANN)

```



## Composition Rasters
Load North America rasters classified for composition and configuration metrics. These rasters are classified by 17 simple land cover classifications.
```{r}
NA_comp_rasters <- list(
  "2012" = rast("Rasters/NA_comp_2012.tif"),
  "2013" = rast("Rasters/NA_comp_2013.tif"),
  "2014" = rast("Rasters/NA_comp_2014.tif"),
  "2015" = rast("Rasters/NA_comp_2015.tif"),
  "2016" = rast("Rasters/NA_comp_2016.tif"),
  "2017" = rast("Rasters/NA_comp_2017.tif"),
  "2018" = rast("Rasters/NA_comp_2018.tif"),
  "2019" = rast("Rasters/NA_comp_2019.tif"),
  "2020" = rast("Rasters/NA_comp_2020.tif"),
  "2021" = rast("Rasters/NA_comp_2021.tif"),
  "2022" = rast("Rasters/NA_comp_2022.tif")
)

```


Create table of all classifications and their respective labels
```{r}
lc_classes <- data.frame(
  class = 1:17,
  label = c("Cereal", "Pulses", "Canola", "Soybean", "Oilseed", "Potatoes", 
            "Berries", "Fruits and Vegetables", "Forage", "Other Ag", 
            "Grassland", "Shrubland", "Wetland", "Forested", "Barren", 
            "Urban", "Water"),
  stringsAsFactors = FALSE)

lc_classes
```


```{r}
utrends_buff<- st_read("Trends_sf/utrends_buff.gpkg", layer = "ubuff")
```


```{r, include = FALSE}

# Initialize an empty list to store results
lsm_NA <- list()

# Loop through each year and buffer, this will result in a very large list. Takes ~6 hours to run

# This worked when only running with shdi and pland. I need to add the additional variables and run. Should take ~8 hours

for (year in names(NA_comp_rasters)) {
  for (i in seq_len(nrow(utrends_buff))) {
    # Transform the buffer to match the raster's CRS
    buffer_i <- st_transform(utrends_buff[i, ], crs = crs(NA_comp_rasters[[year]]))
    
    # Crop and mask the landcover raster for the current year
    class_metrics <- crop(NA_comp_rasters[[year]], buffer_i) |> 
      mask(buffer_i) |> 
      # Calculate class-level metrics
      calculate_lsm(level = "class", metric = c("pland", "lpi")) |> 
      # Add variables to uniquely identify each point and year
      mutate(srd_id = buffer_i$srd_id, year = as.integer(year))
    
    landscape_metrics <- crop(NA_comp_rasters[[year]], buffer_i) |> 
      mask(buffer_i) |> 
      # Calculate landscape-level metrics
      calculate_lsm(level = "landscape", metric = c("shdi", "ed")) |> 
      # Add variables to uniquely identify each point and year
      mutate(srd_id = buffer_i$srd_id, year = as.integer(year))
    
    # Combine class-level and landscape-level metrics
    lsm_NA[[paste(year, i, sep = "_")]] <- bind_rows(class_metrics, landscape_metrics) |> 
      dplyr::select(year, srd_id, class, metric, value)
  }
}

# Combine results into a single data frame
lsm_NA_comp <- bind_rows(lsm_NA)

#save as RDS
#saveRDS(lsm_NA_comp, "Data/phjv_lsm_long.rds")


```

Transform data to wide format.
```{r, eval=FALSE}

# Updated code to transform to wide format
NA_comp <- lsm_NA_comp |> 
  # Fill missing combinations with zeros
  complete(
    year,                        # Include all years
    srd_id,                      # Ensure all buffers are represented
    class = unique(lc_classes$class), # Ensure all classes are included
    metric = c("pland", "lpi"), # Include all metrics
    fill = list(value = 0)       # Fill missing values with zeros
  ) |> 
  # Join with descriptive class labels
  left_join(dplyr::select(lc_classes, class, label), by = "class") |> 
  # Transform to wide format
  pivot_wider(
    values_from = value,                          # Metric values to populate wide-format table
    names_from = c(class, label, metric),   # Columns formed by year, class, label, and metric
    names_glue = "{metric}_c{str_pad(class, 2, pad = '0')}_{label}", # Format names
    names_sort = TRUE,                            # Ensure logical sorting of columns
    values_fill = 0                               # Fill missing combinations with zeros
  ) |> 
  arrange(srd_id)                           # Sort rows by year and buffer ID

NA_comp <- NA_comp[, -c(3,4, 40,41)]
```

Save output landscape metrics
```{r, eval=FALSE}
#save as RDS
saveRDS(NA_comp, "Data/NA_comp.rds")

#save a CSV 
write.csv(NA_comp, "Data/NA_comp.csv", row.names = FALSE)

```

## Landscape Simplification Rasters
Load North America rasters classified for landscape simplification metrics. These rasters are classified by 3 categories
```{r}
NA_simp_rasters <- list(
  "2012" = rast("Rasters/NA_simp_2012.tif"),
  "2013" = rast("Rasters/NA_simp_2013.tif"),
  "2014" = rast("Rasters/NA_simp_2014.tif"),
  "2015" = rast("Rasters/NA_simp_2015.tif"),
  "2016" = rast("Rasters/NA_simp_2016.tif"),
  "2017" = rast("Rasters/NA_simp_2017.tif"),
  "2018" = rast("Rasters/NA_simp_2018.tif"),
  "2019" = rast("Rasters/NA_simp_2019.tif"),
  "2020" = rast("Rasters/NA_simp_2020.tif"),
  "2021" = rast("Rasters/NA_simp_2021.tif"),
  "2022" = rast("Rasters/NA_simp_2022.tif")
)

```

label classifications
```{r}
simp_classes <- data.frame(
  class = 0:2,
  label = c("IR", "Semi_nat", "Ag"),
  stringsAsFactors = FALSE)

simp_classes
```


Run landscape metrics
```{r,  include = FALSE}
lsm_simp <- list()

# Loop through each year in the raster list
for (year in names(NA_simp_rasters)) {
  for (i in seq_len(nrow(utrends_buff))) {
    buffer_i <- st_transform(utrends_buff[i, ], crs = crs(NA_simp_rasters[[year]]))
    
    # Crop and mask the landcover raster for the current year
    metrics <- crop(NA_simp_rasters[[year]], buffer_i) |> 
      mask(buffer_i) |> 
      # Calculate landscape metrics
      calculate_lsm(level = "class", metric = c("pland")) |> 
      # Add variables to uniquely identify each point and year
      mutate(srd_id = buffer_i$srd_id, year = as.integer(year)) |> 
      dplyr::select(year, srd_id, class, metric, value)
    
    # Store the result in the list
    lsm_simp[[paste(year, i, sep = "_")]] <- metrics
  }
}

# Combine results into a single data frame
lsm_NA_simp <- bind_rows(lsm_simp)

# View the combined results
print(lsm_NA_simp)

```


```{r, eval=FALSE}

# Updated code to transform to wide format
NA_simp <- lsm_NA_simp |> 
  # Fill missing combinations with zeros
  complete(
    year,                        # Include all years
    srd_id,                      # Ensure all buffers are represented
    class = unique(simp_classes$class), # Ensure all classes are included
    metric = c("pland"), # Include all metrics
    fill = list(value = 0)       # Fill missing values with zeros
  ) |> 
  # Join with descriptive class labels
  left_join(dplyr::select(simp_classes, class, label), by = "class") |> 
  # Transform to wide format
  pivot_wider(
    values_from = value,                          # Metric values to populate wide-format table
    names_from = c(class, label, metric),   # Columns formed by year, class, label, and metric
    names_glue = "{metric}_c{str_pad(class, 2, pad = '0')}_{label}", # Format names
    names_sort = TRUE,                            # Ensure logical sorting of columns
    values_fill = 0                               # Fill missing combinations with zeros
  ) |> 
  arrange(srd_id)                           # Sort rows by year and buffer ID

NA_simp <- NA_simp[, -c(3)]
```


Save output Simplification landscape metrics
```{r, eval=FALSE}
#save as RDS
saveRDS(NA_simp, "Data/NA_simp.rds")

#save a CSV 
write.csv(NA_comp, "Data/NA_simp.csv", row.names = FALSE)

```



## Combine data sets


Load North America Landscape Simplification and composition metrics
```{r}
NA_simp <- readRDS("Data/NA_simp.rds")
NA_comp <- readRDS("Data/NA_comp.rds")

```

Combine simplification and composition df to make a single df of all landscape metrics
```{r}
raw_all_lsm <- NA_comp %>%   
  left_join(NA_simp, by = c("srd_id", "year"))
```


```{r}
#save as RDS
saveRDS(raw_all_lsm, "Data/raw_all_lsm.rds")

#save a CSV 
write.csv(raw_all_lsm, "Data/raw_all_lsm.csv", row.names = FALSE)
```




```{r, include=FALSE}
utrends <- readRDS("Data/utrends.rds")
utrends <- utrends[, -c(1:6, 14)]
all_lsm <- readRDS("Data/raw_all_lsm.rds")
```


### SHDI
```{r}
#Slope, Intercept, Mean
shdi_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    shdi_slope = coef(lm(shdi_cNA_NA ~ year))[2],  
    shdi_intercept = coef(lm(shdi_cNA_NA ~ year))[1], 
    shdi_mean = mean(shdi_cNA_NA, na.rm = TRUE)
  )

utrends <- shdi_slopes %>%
  left_join(utrends, by = c("srd_id"))
```



### Landscape Simplification
```{r}
#simplification slope, intercept, and mean
simp_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    simp_slope = coef(lm(pland_c02_Ag ~ year))[2],  
    simp_intercept = coef(lm(pland_c02_Ag ~ year))[1],
    simp_mean = mean(pland_c02_Ag, na.rm = TRUE)
  )

utrends <- simp_slopes %>%
  left_join(utrends, by = c("srd_id"))
```


```{r}
#seminat slope, intercept, and mean
snat_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    snat_slope = coef(lm(pland_c01_Semi_nat ~ year))[2],  
    snat_intercept = coef(lm(pland_c01_Semi_nat ~ year))[1],
    snat_mean = mean(pland_c01_Semi_nat, na.rm = TRUE)
  )

utrends <- snat_slopes %>%
  left_join(utrends, by = c("srd_id"))
```

### Grassland LPI
```{r}
#Grassland LPI slope, intercept, and mean
GLPI_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    GLPI_slope = coef(lm(lpi_c11_Grassland ~ year))[2],  
    GLPI_intercept = coef(lm(lpi_c11_Grassland ~ year))[1],
    GLPI_mean = mean(lpi_c11_Grassland, na.rm = TRUE)
  )
utrends <- GLPI_slopes %>%
  left_join(utrends, by = c("srd_id"))

```


### Forage LPI
```{r}
#Forage LPI slope, intercept, and mean
FLPI_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    FLPI_slope = coef(lm(lpi_c09_Forage ~ year))[2],  
    FLPI_intercept = coef(lm(lpi_c09_Forage ~ year))[1],
    FLPI_mean = mean(lpi_c09_Forage, na.rm = TRUE)
  )
utrends <- FLPI_slopes %>%
  left_join(utrends, by = c("srd_id"))

```


### Edge Density
```{r}
#Edge Density slope intercept and mean
ED_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    ED_slope = coef(lm(ed_cNA_NA ~ year))[2],  
    ED_intercept = coef(lm(ed_cNA_NA ~ year))[1],
    ED_mean = mean(ed_cNA_NA, na.rm = TRUE)
  )
utrends <- ED_slopes %>%
  left_join(utrends, by = c("srd_id"))

```



### Grassland Composition
```{r}
#Grassland composition slope intercept and mean
GR_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    GR_slope = coef(lm(pland_c11_Grassland ~ year))[2],  
    GR_intercept = coef(lm(pland_c11_Grassland ~ year))[1],
    GR_mean = mean(pland_c11_Grassland, na.rm = TRUE)
  )
utrends <- GR_slopes %>%
  left_join(utrends, by = c("srd_id"))

```


### Forage Composition
```{r}
#Slope
FO_slopes <- all_lsm %>%
  group_by(srd_id) %>%
  summarize(
    FO_slope = coef(lm(pland_c09_Forage ~ year))[2],  
    FO_intercept = coef(lm(pland_c09_Forage ~ year))[1],
    FO_mean = mean(pland_c09_Forage, na.rm = TRUE)
  )
utrends <- FO_slopes %>%
  left_join(utrends, by = c("srd_id"))

```

# Periphery 
Load LISA results from GeoDA
```{r}
LISA <- read.csv("C:/Users/Andres/OneDrive - University of Saskatchewan/02_Analysis/Chapter1/eBird trend data/vector_pixels/Moran/LISA_results_all.csv")
LISA <- LISA[,-c(1:6, 8:18, 20)]

LISA$Core <- ifelse(LISA$LISA_CL == 1, 1, 0)

#r bind
utrend <- LISA %>%
  dplyr::select(srd_id, Core) %>%
  left_join(utrends, by = c("srd_id"))

utrends <- LISA %>%
  filter(srd_id %in% utrends$srd_id) %>%  # Retain only locations present in utrends_buff
  dplyr::select(srd_id, Core) %>%  # Keep only relevant columns
  left_join(utrends, by = "srd_id")
```


Subset the points that are core and calculate nearest neighbor
```{r}
core_locations <- utrends[utrends$Core == 1, c("latitude", "longitude")]

nn_result <- nn2(data = core_locations, query = utrends[, c("latitude", "longitude")], k = 1)

utrends$Core_distance <- as.numeric(nn_result$nn.dists)
```




## Save
```{r}
NA_model <- utrends
```


```{r, eval=FALSE}
#save as RDS
saveRDS(NA_model, "Data/NA_model.rds")

#save a CSV 
write.csv(NA_model, "Data/NA_model.csv", row.names = FALSE)

```




Read trend 100 ensemble estimates 
```{r}
utrend_folds <- readRDS("Data/utrend_folds.rds")
utrend_folds <- utrend_folds %>%
    filter(!(srd_id >= 126707 & srd_id <= 171704))

```

Calculate cumulative trend in relative abundance using compound interest formula
```{r}
utrend_folds <- utrend_folds %>%
  mutate(abd_ppy = round(abd_ppy, 7)) %>%
  mutate(abd_trend = 100 * ((1 + abd_ppy / 100)^(2022 - 2012) - 1))
```


Copy NA_model df and remove trend information, leaving srd_id and landcover information
```{r}
srd_data <- NA_model %>%
  dplyr::select(-c(27:35))
```


many to one rbind to combine information for each srd_id to the 100 fold estimates
```{r}
NA_folds <- utrend_folds %>%
  left_join(srd_data, by = "srd_id")

NA_folds <- NA_folds %>%
  dplyr::select(-c(1:2))
```


save NA_folds df
```{r}
saveRDS(NA_folds, "Data/NA_folds.rds")
```
