---
title: "rangewide-lsm"
author: "Andres N. Rosales"
date: "2025-01-13"
output:
  html_document:
    highlight: espresso
    toc: yes
    theme: sandstone
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

# Library
```{r, message=FALSE, warning=FALSE}

library(dplyr)
library(purrr)
library(ggplot2)
library(rnaturalearth)
library(sf)
library(raster)
library(terra)
library(tinytex)
library(rmarkdown)
library(knitr)
library(ebirdst)
library(exactextractr)
library(landscapemetrics)
library(readr)
library(stringr)
library(tidyr)
library(units)
library(viridis)
library(progress)
library(dplyr)

```


Create an object of Upland Sandpiper trends (2012-2022). 
```{r}

utrends_st <- load_trends("Upland Sandpiper")

```

#Spatial Buffer for predictor grid
Removed Alaska / Yukon points from utrends_st
```{r}

utrends_st <- utrends_st %>% filter(!(srd_id >= 126707 & srd_id <= 171704))

```

Calculate annual abundance

$Ay = A2017 × (1 + r)^{y-2017}$

- Ay is the abundance in year y.
- A2017  is the abundance in 2017 (median year).
- r is the annual percent change expressed as a decimal.
- y is the year of interest.

```{r}
#annual abundance estimates
utrends_st <- utrends_st %>%
  bind_cols(map_dfc(2012:2022, 
                    ~ tibble(!!paste0( .x) := utrends_st$abd * 
                               (1 + utrends_st$abd_ppy) ^ (.x - 2017))))
#remove abd column since this is only the abd for the median year
utrends_st <- utrends_st[, -c(4)]


```


Subset estimates with >80% confidence
```{r}
utrends_st <- subset(utrends_st, abd_ppy_nonzero == "TRUE")

```


Using utrend_st make an 'sf' object. Set crs as WGS84 then reproject to NAD83 (5070)
```{r}
#set to 4326 then trnasfomr later during lsm
utrends_buff <- utrends_st |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) |>
  st_buffer(dist = set_units(13.5, "km"))                                        # Create 13.5 km buffers

ggplot() +
  geom_sf(data = utrends_buff, fill = "lightblue", color = "darkblue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "13.5 km Circular Buffers", x = "Longitude", y = "Latitude")


write_sf(utrends_buff,"C:/Users/Andres/OneDrive - University of Saskatchewan/03_Arcpro/Chapter1/Shapefiles/ebirdst/Buffer/utrends_buff.gpkg",
         layer = "ubuff")

```


## Composition Rasters
Load North America rasters classified for composition and configuration metrics. These rasters are classified by 17 simple land cover classifications.
```{r}
NA_comp_rasters <- list(
  "2012" = rast("Rasters/NA_comp_2012.tif"),
  "2013" = rast("Rasters/NA_comp_2013.tif"),
  "2014" = rast("Rasters/NA_comp_2014.tif"),
  "2015" = rast("Rasters/NA_comp_2015.tif"),
  "2016" = rast("Rasters/NA_comp_2016.tif"),
  "2017" = rast("Rasters/NA_comp_2017.tif"),
  "2018" = rast("Rasters/NA_comp_2018.tif"),
  "2019" = rast("Rasters/NA_comp_2019.tif"),
  "2020" = rast("Rasters/NA_comp_2020.tif"),
  "2021" = rast("Rasters/NA_comp_2021.tif"),
  "2022" = rast("Rasters/NA_comp_2022.tif")
)

```


```{r}
check <- rast("Rasters/NA_comp_2012.tif")

check_landscape(check)
```


Create table of all classifications and their respective labels
```{r}
lc_classes <- data.frame(
  class = 1:17,
  label = c("Cereal", "Pulses", "Canola", "Soybean", "Oilseed", "Potatoes", 
            "Berries", "Fruits and Vegetables", "Forage", "Other Ag", 
            "Grassland", "Shrubland", "Wetland", "Forested", "Barren", 
            "Urban", "Water"),
  stringsAsFactors = FALSE)

lc_classes
```

```{r, eval=FALSE}

# Initialize an empty list to store results
lsm_NA <- list()

# Loop through each year and buffer, this will result in a very large list. Takes ~6 hours to run

# This worked when only running with shdi and pland. I need to add the additional variables and run. Should take ~8 hours

for (year in names(NA_comp_rasters)) {
  for (i in seq_len(nrow(utrends_buff))) {
    # Transform the buffer to match the raster's CRS
    buffer_i <- st_transform(utrends_buff[i, ], crs = crs(NA_comp_rasters[[year]]))
    
    # Crop and mask the landcover raster for the current year
    class_metrics <- crop(NA_comp_rasters[[year]], buffer_i) |> 
      mask(buffer_i) |> 
      # Calculate class-level metrics
      calculate_lsm(level = "class", metric = c("pland", "lpi")) |> 
      # Add variables to uniquely identify each point and year
      mutate(srd_id = buffer_i$srd_id, year = as.integer(year))
    
    landscape_metrics <- crop(NA_comp_rasters[[year]], buffer_i) |> 
      mask(buffer_i) |> 
      # Calculate landscape-level metrics
      calculate_lsm(level = "landscape", metric = c("shdi", "ed")) |> 
      # Add variables to uniquely identify each point and year
      mutate(srd_id = buffer_i$srd_id, year = as.integer(year))
    
    # Combine class-level and landscape-level metrics
    lsm_NA[[paste(year, i, sep = "_")]] <- bind_rows(class_metrics, landscape_metrics) |> 
      dplyr::select(year, srd_id, class, metric, value)
  }
}

# Combine results into a single data frame
lsm_NA_comp <- bind_rows(lsm_NA)

#save as RDS
#saveRDS(lsm_NA_comp, "Data/phjv_lsm_long.rds")


```

Transform data to wide format.
```{r, eval=FALSE}

# Updated code to transform to wide format
NA_comp <- lsm_NA_comp |> 
  # Fill missing combinations with zeros
  complete(
    year,                        # Include all years
    srd_id,                      # Ensure all buffers are represented
    class = unique(lc_classes$class), # Ensure all classes are included
    metric = c("pland", "lpi"), # Include all metrics
    fill = list(value = 0)       # Fill missing values with zeros
  ) |> 
  # Join with descriptive class labels
  left_join(dplyr::select(lc_classes, class, label), by = "class") |> 
  # Transform to wide format
  pivot_wider(
    values_from = value,                          # Metric values to populate wide-format table
    names_from = c(class, label, metric),   # Columns formed by year, class, label, and metric
    names_glue = "{metric}_c{str_pad(class, 2, pad = '0')}_{label}", # Format names
    names_sort = TRUE,                            # Ensure logical sorting of columns
    values_fill = 0                               # Fill missing combinations with zeros
  ) |> 
  arrange(srd_id)                           # Sort rows by year and buffer ID

```

Save output landscape metrics
```{r, eval=FALSE}
#save as RDS
saveRDS(NA_comp, "Data/NA_comp.rds")

#save a CSV 
write.csv(NA_comp, "Data/NA_comp.csv", row.names = FALSE)

```

NA_comp = DF with final landscape metrics for rangewide analysis. At this point it does not have trend data added to the DF

Load phjv Composition metrics
```{r}
NA_comp <- readRDS("Data/NA_comp.rds")
```


## Landscape Simplification Rasters
Load North America rasters classified for landscape simplification metrics. These rasters are classified by 3 categories
```{r}
NA_simp_rasters <- list(
  "2012" = rast("Rasters/NA_simp_2012.tif"),
  "2013" = rast("Rasters/NA_simp_2013.tif"),
  "2014" = rast("Rasters/NA_simp_2014.tif"),
  "2015" = rast("Rasters/NA_simp_2015.tif"),
  "2016" = rast("Rasters/NA_simp_2016.tif"),
  "2017" = rast("Rasters/NA_simp_2017.tif"),
  "2018" = rast("Rasters/NA_simp_2018.tif"),
  "2019" = rast("Rasters/NA_simp_2019.tif"),
  "2020" = rast("Rasters/NA_simp_2020.tif"),
  "2021" = rast("Rasters/NA_simp_2021.tif"),
  "2022" = rast("Rasters/NA_simp_2022.tif")
)

```

label classifications
```{r}
simp_classes <- data.frame(
  class = 0:2,
  label = c("IR", "Semi_nat", "Ag"),
  stringsAsFactors = FALSE)

simp_classes
```


Run landscape metrics
```{r, eval=FALSE}
lsm_simp <- list()

# Loop through each year in the raster list
for (year in names(NA_simp_rasters)) {
  for (i in seq_len(nrow(utrends_buff))) {
    buffer_i <- st_transform(utrends_buff[i, ], crs = crs(NA_simp_rasters[[year]]))
    
    # Crop and mask the landcover raster for the current year
    metrics <- crop(NA_simp_rasters[[year]], buffer_i) |> 
      mask(buffer_i) |> 
      # Calculate landscape metrics
      calculate_lsm(level = "class", metric = c("pland")) |> 
      # Add variables to uniquely identify each point and year
      mutate(srd_id = buffer_i$srd_id, year = as.integer(year)) |> 
      dplyr::select(year, srd_id, class, metric, value)
    
    # Store the result in the list
    lsm_simp[[paste(year, i, sep = "_")]] <- metrics
  }
}

# Combine results into a single data frame
lsm_NA_simp <- bind_rows(lsm_simp)

# View the combined results
print(lsm_NA_simp)

```


```{r, eval=FALSE}

# Updated code to transform to wide format
NA_simp <- lsm_NA_simp |> 
  # Fill missing combinations with zeros
  complete(
    year,                        # Include all years
    srd_id,                      # Ensure all buffers are represented
    class = unique(simp_classes$class), # Ensure all classes are included
    metric = c("pland"), # Include all metrics
    fill = list(value = 0)       # Fill missing values with zeros
  ) |> 
  # Join with descriptive class labels
  left_join(dplyr::select(simp_classes, class, label), by = "class") |> 
  # Transform to wide format
  pivot_wider(
    values_from = value,                          # Metric values to populate wide-format table
    names_from = c(class, label, metric),   # Columns formed by year, class, label, and metric
    names_glue = "{metric}_c{str_pad(class, 2, pad = '0')}_{label}", # Format names
    names_sort = TRUE,                            # Ensure logical sorting of columns
    values_fill = 0                               # Fill missing combinations with zeros
  ) |> 
  arrange(srd_id)                           # Sort rows by year and buffer ID

```


Save output Simplification landscape metrics
```{r, eval=FALSE}
#save as RDS
saveRDS(NA_simp, "Data/NA_simp.rds")

#save a CSV 
write.csv(NA_comp, "Data/NA_simp.csv", row.names = FALSE)

```


Load North America Landscape Simplification metrics
```{r}
phjv_lsm_simp <- readRDS("Data/NA_simp.rds")
```
