---
title: "Rangewide Analysis"
author: "Andres N. Rosales"
date: "2025-01-24"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---

# Candidate Models

**Prediction 1**

Breeding areas that increased in abundance will be in more diverse landscapes with decreased (or low) agricultural intensity.
Landscapes that became simplified agriculturally over the period will have decreased abundance.

$Trend = Δ SHDI +  mean SHDI + Δ Edge Density + mean Edge Density + mean Simplification + Δ Simplification$

Breeding areas peripheral to historical core breeding areas experienced declines driven by homogeneous landscapes.

$Trend = Δ SHDI + mean SHDI + Δ Edge Density + mean Edge Density + mean Simplification + Δ Simplification$


**Prediction 2**

Greater landscape heterogeneity will lower the negative effects of landscape simplification on breeding trends. 
“Is the effect of landscape diversity on Upland Sandpiper population dynamics dependent on the degree of simplification?”

$Trend = Δ SHDI + Δ Cropland + Δ SHDI * Δ Cropland $

**Prediction 3**

Forage agriculture and grasslands are associated with positive trends and high abundance. 
“How are trends in forage agriculture and grasslands associated with trends in abundance of Upland Sandpiper?”

$Trend = Δ Forage Ag + Δ Grassland$



**Prediction 4**

Trends decline or become more variable with increasing distance from the core breeding range. Conversely, trends will be more stable near the core breeding region.

$Trend = Periphery $

**Prediction 5**

Geographical variation in abundance is driven by marginal habitat at the periphery.

$Trend = Periphery + Δ SHDI +  Δ Cropland$

# Library

```{r, message=FALSE, warning=FALSE, include=FALSE}

library(dplyr)
library(purrr)
library(sf)
library(ggplot2)
library(tinytex)
library(rmarkdown)
library(knitr)
library(tidyr)
library(units)
library(dplyr)
library(lme4)
library(mgcv)
library(GGally)
library(car)
library(gratia)
library(ggcorrplot)
library(brms)
library(tidybayes)
library(cmdstanr)
library(GGally)
```

```{r, include=FALSE}
#Edited trend df
NA_model <- readRDS("Data/NA_model.rds")
names(NA_model)
```


```{r}
NA_model <- NA_model %>%
  mutate(CI_width = abd_trend_upper - abd_trend_lower)

NA_model <- NA_model %>% 
  filter(CI_width <= 35)
```

# Response Distribution

Distribution of trend response variable.

```{r, echo = FALSE }
ggplot(NA_model, aes(x = abd_trend)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  theme_minimal()

```

```{r, echo=FALSE}
#Distribution of abundance response variable. 
ggplot(NA_model, aes(x = abd)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  theme_minimal()

```

## Periphery

Mapping of trend estimates with red points identified as "core"breeding areas as defined by Morans I.

```{r, echo=FALSE}
core_map <- NA_model |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) |>
  st_buffer(dist = set_units(13.5, "km")) 

ggplot() +
  geom_sf(data = core_map, aes(fill = factor(Core)), color = "black", size = 0.1, alpha = 1) +
  scale_fill_manual(values = c("0" = "white", "1" = "red")) +
  theme_minimal() +
  labs(title = "13.5 km Circular Buffers", x = "Longitude", y = "Latitude")

```

Distribution of abundance estimates.

# Predictor relationships

Correlation matrix before aggregating predictors.

```{r, echo=FALSE}
# List of predictors to include
mean_predictors <- c(
    "shdi_mean" , "simp_mean" ,  "ED_mean" , "FO_mean" , "abd" , "Core_distance", "snat_mean" 
)

mean_filtered_data <- NA_model %>%
  dplyr::select(all_of(mean_predictors))

mean_cor_matrix <- cor(mean_filtered_data, use = "complete.obs", method = "pearson")

ggcorrplot(mean_cor_matrix, lab = TRUE, lab_size = 3, colors = c("red", "white", "blue"))
```

I still need to determine a pearsons correlation coefficient threshold to report. Some literature has used 0.7 and I will go with that for now. I still need to develop a better backing for this and if this is something I will continue to use going forward...

```{r}
# List of predictors to include
slope_predictors <- c(
    "shdi_slope" , "simp_slope" , "ED_slope" , "FO_slope" , "GR_slope" , "Core_distance", "snat_slope" 
)

slope_filtered_data <- NA_model %>%
  dplyr::select(all_of(slope_predictors))

slope_cor_matrix <- cor(slope_filtered_data, use = "complete.obs", method = "pearson")

ggcorrplot(slope_cor_matrix, lab = TRUE, lab_size = 3, colors = c("red", "white", "blue"))
```

```{r}
# List of predictors to include
all_predictors <- c(
    "shdi_mean" , "simp_mean" ,  "ED_mean" , "FO_mean" , "GR_mean" , "abd", "Core_distance", "snat_mean", "shdi_slope" , "simp_slope" ,  "ED_slope" , "FO_slope" , "GR_slope" , "snat_slope" 
)

all_filtered_data <- NA_model %>%
  dplyr::select(all_of(all_predictors))

all_cor_matrix <- cor(all_filtered_data, use = "complete.obs", method = "pearson")

ggcorrplot(all_cor_matrix, lab = TRUE, lab_size = 3, colors = c("red", "white", "blue"))
```

```{r,  echo=FALSE ,message=FALSE, fig.width=10, fig.height=7}

# Create ggpairs plot
ggpairs(
  all_filtered_data,
  lower = list(continuous = wrap("smooth", colour = "blue", size = 1)),
  diag = list(continuous = "densityDiag"), 
  upper = list(continuous = "cor") 
)
```

## Shannon Diversity Index

```{r, echo=FALSE}
ggplot(NA_model, aes(x = shdi_mean, y = abd_trend)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "loess",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between SHDI Slope and Trend",
    x = "Slope of SHDI (Change over Time)",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

```

## Landscape Simplification

plot simplification slope against trend

```{r, echo=FALSE}
ggplot(NA_model, aes(x = simp_mean, y = abd_trend)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "loess",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Simplification Slope and Trend",
    x = "Mean Simplification",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

```{r}
SI_cube <- glm(abd_trend ~ simp_mean + I(simp_mean^2) + I(simp_mean^3), data = NA_model)
summary(SI_cube)

SI_lin <- glm(abd_trend ~ simp_mean, data = NA_model)
summary(SI_lin)

AIC(SI_cube, SI_lin)
```

Plotting of linear and quad model for mean landscape simplification.

```{r, echo=FALSE}
SI_seq<-seq(0,90,25)
SIQ <- predict(SI_cube,list(simp_mean=SI_seq),type="response")
plot(NA_model$simp_mean, NA_model$abd_trend, pch=16,cex.axis=1.6,cex.lab=1.6,cex=0.5,col="red", xlab = "Mean Simplification", ylab = "Trend" )
abline(SI_lin, lwd = 2, col = "blue")
lines(SI_seq,SIQ,lwd=2)
```

## LPI Grassland

Plot trend \~ slope and trend \~ Mean

```{r, echo=FALSE}
ggplot(NA_model, aes(x = GLPI_mean, y = abd_trend)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "loess", se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Grassland LPI and Trend",
    x = "Mean Grassland LPI",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

```

## LPI Forage

Plot trend \~ slope and trend \~ Mean

```{r, echo=FALSE}
ggplot(NA_model, aes(x = log(FLPI_mean), y = abd_trend)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "loess",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Forage LPI and Trend",
    x = "Mean forage LPI",
    y = "Trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

```

## Edge Density

calculate the slope and average LPI for each location

```{r, echo=FALSE}
ggplot(NA_model, aes(x = ED_mean, y = abd_trend)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "loess",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between Edge Density and Trend",
    x = "Mean Edge Density",
    y = "Abundance Trend (abd_trend)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

```{r}
ED_quad <- glm(abd_trend ~ ED_mean + I(ED_mean^2), data = NA_model)
summary(ED_quad)
ED_lin <- glm(abd_trend ~ ED_mean , data = NA_model)
summary(ED_lin)
AIC(ED_quad, ED_lin)
```

Plot of linear and quad model for Mean edge density.

```{r, echo=FALSE}
ED_seq<-seq(0,250,25)
EDQ <- predict(ED_quad,list(ED_mean=ED_seq),type="response")
plot(NA_model$ED_mean, NA_model$abd_trend, pch=16,cex.axis=1.6,cex.lab=1.6,cex=0.5,col="red", xlab = "Mean Edge Density", ylab = "Trend" )
abline(ED_lin, lwd = 2, col = "blue")
lines(ED_seq,EDQ,lwd=2)
```

### Periphery

```{r, echo=FALSE}
ggplot(NA_model, aes(x = log(Core_distance), y = abd_trend)) +
  geom_point( alpha = 0.3, size = 2) +  
  geom_smooth(method = "loess",  se = TRUE, linetype = "dashed") + 
  labs(
    title = "Relationship Between distance from core and Trend",
    x = "Distance from Core Breeding",
    y = "Abundance Trend (abd_trend)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

# Model Runs

## Global Models

Removed both LPI metrics for now due to high correlation with many variables.


### General Additive Model

GAM 1

```{r}
gam_global <- gam(abd_trend ~ 
                    s(FO_mean) + 
                    s(GR_mean) + 
                    s(ED_mean) +  
                    s(shdi_mean) + 
                    s(simp_mean) + 
                    (abd) +  
                    te(shdi_mean, simp_mean) + 
                    s(longitude, latitude, bs= "gp", k = 300 ) + 
                    s(Core_distance), 
                  data = NA_model, method = "ML")
summary(gam_global)
draw(gam_global)
gam.check(gam_global)

```

GAM 2 Based on the EDF in the gam_global model, I will change terms with EDF near 1 to linear terms and drop non-informative parameters

```{r}
gam_global2 <- gam(abd_trend ~ 
                     s(FO_mean) + 
                     s(GR_mean) + 
                     s(ED_mean) +  
                     s(shdi_mean) + 
                     simp_mean + 
                     s(abd) +  
                     te(shdi_mean, simp_mean) +
                     s(longitude, latitude), data = NA_model, method = "ML")
summary(gam_global2)
gam.check(gam_global2)
draw(gam_global2)
```

Substitute the smooth term for location for the fixed term of distance from core. distance was significant in a single model but not in the above gams because the smoothed term of location is likely absorbing much of its variation

```{r}
gam_global3 <- gam(abd_trend ~
                     s(FO_mean) + 
                     s(GR_mean) + 
                     s(ED_mean, k = 30) +  
                     shdi_mean + 
                     simp_mean + 
                     s(abd) +  
                     te(shdi_mean, simp_mean) + 
                     s(longitude, latitude, bs= "tp", k = 300), 
                   data = NA_model, method = "ML")
summary(gam_global3)
draw(gam_global3)
gam.check(gam_global3)
```

```{r}
gam_global4 <-  gam(abd_trend ~
                     s(FO_mean) + 
                     s(FO_slope) +
                     s(GR_mean) + 
                     s(GR_slope) +
                     s(ED_mean, k = 30) + 
                     s(ED_slope) +
                     shdi_mean + 
                     s(shdi_slope) +
                     simp_mean + 
                     s(simp_slope) +
                     s(abd) +  
                     te(shdi_mean, simp_mean) + 
                     te(shdi_slope, simp_mean) + 
                     te(shdi_mean, shdi_slope) +
                     te(simp_mean, simp_slope) +
                     s(longitude, latitude, bs= "tp", k = 300), 
                   data = NA_model, method = "REML")

summary(gam_global4)
gam.check(gam_global4)
draw(gam_global4)
```



### Prediction 1

```{r}
P1gam <- gam(abd_trend ~ s(shdi_slope) + 
               s(shdi_mean) + 
               s(ED_slope, k = 35) + 
               s(ED_mean, k = 35) + 
               simp_slope + simp_mean + 
               s(longitude, latitude, bs = "tp", k = 500), 
             data = NA_model)
summary(P1gam)
gam.check(P1gam)
plot(P1gam, pages = 1, seWithMean = TRUE)
```

```{r}
P1gam2 <- gam(abd_trend ~ 
                s(shdi_mean) + 
                s(ED_mean, k = 30) + 
                s(snat_mean) + 
                s(longitude, latitude, bs= "tp", k = 300), 
              data = NA_model)
summary(P1gam2)
plot(P1gam2, pages = 1, seWithMean = TRUE)
gam.check(P1gam2)
```

### Prediction 2

```{r}
P2gam <- gam(abd_trend ~ s(shdi_mean) + 
               s(simp_mean) + 
               te(shdi_mean, simp_mean) + 
               s(longitude, latitude, bs= "tp", k = 300)
             , data = NA_model)
summary(P2gam)

draw(P2gam)
```


```{r}
P2gamslope <- gam(abd_trend ~ s(shdi_slope) + 
               s(simp_slope) + 
               te(shdi_slope, simp_slope) + 
               s(longitude, latitude, bs= "tp", k = 300)
             , data = NA_model)
summary(P2gamslope)
gam.check(P2gamslope)
draw(P2gamslope)
```
### Prediction 3

```{r}
P3gam <- gam(abd_trend ~ FO_slope + 
               s(FO_mean) + 
               GR_slope + 
               s(GR_mean, k = 15) + 
               s(longitude, latitude, bs= "tp", k = 300),
            data = NA_model)
summary(P3gam)
gam.check(P3gam) 
draw(P3gam)
```


### Prediction 4


```{r}
P4gam <- gam(abd_trend ~ Core_distance + 
             s(longitude, latitude, bs = "tp", k = 300),
            data = NA_model)
summary(P4gam)
gam.check(P4gam)
draw(P4gam)
```
### Prediction 5


```{r}
P5gam <- gam(abd_trend ~ te(shdi_slope, Core_distance) + 
               te(simp_slope, Core_distance) + 
               s(longitude, latitude, bs = "tp", k = 300),
            data = NA_model)
summary(P5gam)
gam.check(P5gam)
draw(P5gam)
```


# Bayesian GAM

Read df with all estimates and their 100 folds

```{r, eval=FALSE, include = FALSE}
NA_folds <- readRDS("Data/NA_folds.rds")
```

test run on a small sample

```{r, eval=FALSE, include = FALSE}
# Select 5 random locations
test_locations <- sample(unique(NA_folds$srd_id), 50)

# Subset the data to include only these locations
NA_fol <- NA_folds %>%
  filter(srd_id %in% test_locations)

NA_fol <- NA_folds %>%
 filter(!(srd_id >= 242031))
```

Prior check

```{r, eval=FALSE, include = FALSE}
# Run the model using priors only
prior_check <- brm(
  abd_trend ~ (1 | srd_id),
  data = NA_folds,
  family = gaussian(),
  prior = c(
    prior(normal(-13.98, 12), class = Intercept),  # Our informative prior
    prior(exponential(1), class = sd),
    prior(exponential(1), class = sigma)
  ),
  sample_prior = "only",  # Generate samples only from priors, not the data
  chains = 2,                
  iter = 4000,              
  warmup = 1000,            
  cores = 4                 
)

# Plot the prior predictive distribution for the intercept
plot(prior_check)
```

```{r, eval=FALSE, include = FALSE}
bayes_model1 <- brm(
  abd_trend ~ 
     (1 | srd_id),
  data = NA_folds,
  family = gaussian(),  
  chains = 4,                  # Use 4 chains for better convergence
  iter = 12000,                 # More iterations to improve ESS
  warmup = 2000,               # Longer warmup to stabilize chains
  cores = 8,                   # Fully utilize your 8-core system
  control = list(
    adapt_delta = 0.99,        # More conservative step size to avoid divergences
    max_treedepth = 15         # Allows deeper exploration of the posterior
  )
)

```

```{r, eval=FALSE, include = FALSE}
saveRDS(bayes_model1, "Data/Bayesm1.rds")

summary(bayes_model1)

post_summary <- bayes_model1 %>%
  spread_draws(r_srd_id[srd_id,]) %>%  # Correct structure for random effects
  median_qi(trend_est = r_srd_id)

post_summary <- post_summary %>%
  mutate(ci_width = .upper - .lower)

print(post_summary)

saveRDS(post_summary, "Data/post_summary.rds")
```

```{r, eval=FALSE, include = FALSE}
ggplot(post_summary_test, aes(x = trend_est)) +
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.6, color = "black") +
  theme_minimal() +
  labs(
    title = "Distribution of Median Posterior Trend Estimates",
    x = "Median Posterior Trend Estimate",
    y = "Count (Number of Locations)"
  )


ggplot(post_summary_test, aes(x = reorder(srd_id, trend_est), y = trend_est)) +
  geom_point(color = "blue", size = 2) +  # Plot median estimates
  geom_errorbar(aes(ymin = .lower, ymax = .upper), width = 0.2) +  # Add 95% CIs
  theme_minimal() +
  labs(
    title = "Posterior Trend Estimates per Location",
    x = "Location ID (Ordered by Trend)",
    y = "Median Posterior Trend Estimate"
  ) +
  theme(axis.text.x = element_blank())  # Hide x-axis labels if too many locations


```

## BRM model 2 with priors

```{r, eval=FALSE, include = FALSE}
bayes_model2 <- brm(
  abd_trend ~ (1 | srd_id),  
  data = NA_folds,  
  family = gaussian(),
  
  prior = c(
    prior(normal(-13.98, 12), class = Intercept),  
    prior(exponential(1), class = sd),             
    prior(exponential(1), class = sigma)           
  ),
  chains = 4,                
  iter = 12000,              
  warmup = 2000,             
  cores = 16,  
  control = list(
    adapt_delta = 0.95,      
    max_treedepth = 12       
  )
)

```

```{r, eval=FALSE, include = FALSE}
summary(bayes_model2)
post_summary2 <- bayes_model2 %>%
  spread_draws(r_srd_id[srd_id,]) %>%  
  median_qi(trend_est = r_srd_id)

post_summary2 <- post_summary2 %>%
  mutate(ci_width = .upper - .lower)

print(post_summary2)

saveRDS(post_summary2, "Data/post_summary2.rds")

plot(bayes_model2)
```

```{r, eval=FALSE, include = FALSE}
ggplot(post_summary2, aes(x = ci_width)) +
  geom_histogram(binwidth = 0.005, fill = "blue", alpha = 0.6, color = "black") +
  theme_minimal() +
  labs(
    title = "Distribution of Ci widths",
    x = "CI Width",
    y = "Count (Number of Locations)"
  )

ggplot(post_summary2, aes(x = srd_id, y = ci_width)) +
  geom_point() +
  geom_hline(yintercept = median(post_summary2$ci_width), linetype = "dashed", color = "red") +
  labs(title = "CI Widths Across Locations",
       y = "Credible Interval Width",
       x = "Location (srd_id)") +
  theme_minimal()
```

## Gaussian Process model

Make subset of data 
```{r}
NA_folds <- NA_folds %>%
  filter(srd_id >= 230274 & srd_id <= 240815)
```


```{r, eval=FALSE, include = FALSE}
bayes_model_gp <- brm(
  abd_trend ~ gp(longitude, latitude, k = 5, c = 5/4),  # GP smooths spatial variation
  data = NA_folds,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = Intercept)),
  chains = 4,                
  iter = 4000,              
  warmup = 2000,            
  cores = 4,  
  threads = threading(2),
  backend = "cmdstanr",
  control = list(
    adapt_delta = 0.99,      
    max_treedepth = 8       
  )
)
```

```{r, eval=FALSE, include = FALSE}
summary(bayes_model_gp)
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
cmdstanr::install_cmdstan()
library(cmdstanr)
plot(bayes_model_gp)
```
